Algebraic File Synchronization: Adequacy and Completeness

### 2016-10-15 p2/axioms.tex
Improved text

So that we can describe the effects of commands in general,
let us introduce some notation
and note some observations.
We already know that
commands usually do not occur in isolation,
and are applied to filesystems in time.
Therefore we will investigate sequences of commands with a well-defined order.
\begin{mydef}[Sequences of commands: $\cc$ and $\emptyseq$]
We use ``$\cc$'' to concatenate commands to form a sequence, or concatenate sequences to form a longer sequence,
with the meaning that the command on the left is applied first:
\[ \alpha\cc\beta\aFS = \beta(\alpha\aFS). \]
For example, we will write
$S=\cbdaa{n}\cc\cbfaa{m}$, or $S'=S\cc\cbba{o}$.
% monoid, but not free monoid
% monoid: semigroup with identity element
% Sequences of commands form a free semigroup
Sequences of commands form a monoid, and, as usual,
$\emptyseq$ denotes its unit element, the empty sequence,
which is defined on all filesystems and by definition leaves all filesystems unchanged.
\end{mydef}

\begin{mydef}[$\Dom{S}$ and $\Img{S}$]
For a sequence of commands $S$, $\Dom{S}$ is the domain of $S$, which is the set of filesystems $S$ is defined on,
that is, which it does not break.

@@

The following two relations 
(which echo the ones defined in \cite{NREC})
allow us to investigate sequences independently of filesystems.
In the definitions, $A,B,S,T$ stand for arbitrary sequences of commands.

@@

It is easy to see that equivalence holds on the level of filesystems:

\begin{mylem}
$A\equiv B$
iff $A$ and $B$ behave in the same way on
all filesystems, that is, $\forall \FS: A\aFS=B\aFS$.
\end{mylem}
\begin{proof}
If $\forall \FS: A\aFS=B\aFS$ then $A\eqext B$ and $B\eqext A$ and so by definition $A\equiv B$.
For the reverse statement we assume that $A\eqext B$ and $B\eqext A$.
$B\eqext A$ implies that it is not possible for both $B\aFS\neq\fsbroken$ and $A\aFS=\fsbroken$ to be true.
Therefore, for every $\FS$ either $A\aFS\neq\fsbroken$, in which case $A\aFS=B\aFS$ as $A\eqext B$, or $A\aFS=\fsbroken$,
in which case $B\aFS=\fsbroken$ must be true, and therefore $A\aFS=B\aFS$ still holds.
\end{proof}

@@

We methodically investigate all possibilities using a computer 
program\footnote{The program is accessible on-line at \\
https://github.com/csirmaz/AlgebraicSyncPaper/blob/master/p2/prove.py.}
to determine
which pairs of commands cause errors all the time,
which can be simplified to one command or $\emptyseq$,
and which commute and can be reversed without any change in their overall effect.
Our basis for this investigation was the model of filesystems introduced in this paper.
These properties of command pairs are crucial as they determine
how a set of commands can be re-ordered to be applied to a filesystem
during synchronization, and what commands will never be compatible.
Below we list a number of statements derived using this method.

@@

\begin{mydef}[Assertion command]
A command is an assertion command if
for every filesystem it either breaks it or leaves it in the same state.
In other words, assertion commands have the same input and output type,
and there must also be only one possible value for their output.
Accordingly, $\cbba{n}$ and $\cdda{n}$ are the only two types of assertion commands.
\end{mydef}

@@

\begin{mylem}\label{incomparable_is_independent}
Two different commands that are not assertion commands are independent iff the nodes they change are incomparable:
\[ \cxynv\indep\czwmv \Leftrightarrow n\unrel m \]
if $\cxynv\neq\czwmv$ and none of $\cxy$ and $\cxw$ is $\cbb$ or $\cdd$.
\end{mylem}
\begin{proof}
This is easy to show based on the output of the software that investigates the behavior of command pairs.
However, this proposition can also be derived from the \namecrefs{ax_separate_commute} already listed
in the following way.

We proceed in two steps. First, we note that
from \cref{ax_separate_commute,ax_separate_nobreaks} we know that
if $n\unrel m$, then $\cxynv$ and $\czwmv$ must necessarily be independent, and therefore
$ \cxynv\nindep\czwmv \wedge n\unrel m $ cannot be the case.

Next, we prove that
\[ \cxynv\indep\czwmv \wedge n\nunrel m \] is also impossible as it would lead to contradiction.
As $n\nunrel m$
we know that $n=m$, $n\descendant m$ or $m\descendant n$.

If $n=m$, then from \cref{ax_same_breaks} we know $[y]=[z]$ and $[w]=[x]$
as otherwise the commands, in one order or the other, would break all filesystems
and they could not be independent.
The commands are not assertion commands, and so from \cref{ax_same_singlec}
we know that $\cxynv\cc\czwnv\equiv\czwnv\cc\cxynv$ can only happen if $y=w$.
Therefore $[x]=[y]=[z]=[w]$ which is only possible if both commands are $\cff$ and have the same
output value. This, however, contradicts our assumption that they are different.

>>>>

So that we can describe the effects of commands in general,
let us introduce some notation
and note some observations.
We already know that
commands usually do not occur in isolation,
and are applied to filesystems in time.
Therefore we will investigate sequences of commands with a well-defined order.
\begin{mydef}[Sequences of commands: $\cc$ and $\emptyseq$]
We use ``$\cc$'' to combine commands to form a sequence, or combine sequences to form a longer sequence,
with the meaning that the command on the left is applied first:
\[ \alpha\cc\beta\aFS = \beta(\alpha\aFS). \]
% For example, we will write
% $S=\cbdaa{n}\cc\cbfaa{m}$, or $S'=S\cc\cbba{o}$.
%% monoid, but not free monoid
%% monoid: semigroup with identity element
%% Sequences of commands form a free semigroup
Sequences of commands form a monoid, and, as usual,
$\emptyseq$ denotes its unit element, the empty sequence,
which is defined on all filesystems and by definition leaves all filesystems unchanged.
\end{mydef}

\begin{mydef}[$\Dom{S}$ and $\Img{S}$]
For a sequence of commands $S$, $\Dom{S}$ is the domain of $S$,
which is the set of filesystems $S$
(that is, the commands it contains in the given order)
does not break.

@@

The following two relations 
(which echo the ones defined in \cite{NREC})
allow us to investigate sequences of commands independently of filesystems.
In the definitions, $A,B,S,T$ stand for arbitrary sequences.

@@

It is easy to see that this equivalence holds on the level of filesystems:

\begin{mylem}
$A\equiv B$
iff $A$ and $B$ behave in the same way on
all filesystems, that is, $\forall \FS: A\aFS=B\aFS$.
\end{mylem}
\begin{proof}
If $\forall \FS: A\aFS=B\aFS$ then $A\eqext B$ and $B\eqext A$ and so by definition $A\equiv B$.
For the reverse statement we know $A\eqext B$ and $B\eqext A$.
$B\eqext A$ implies that it is not possible for both $B\aFS\neq\fsbroken$ and $A\aFS=\fsbroken$ to be true.
Therefore, for every filesystem $\FS$ either $A\aFS\neq\fsbroken$, in which case $A\aFS=B\aFS$ as $A\eqext B$, or $A\aFS=\fsbroken$,
in which case $B\aFS=\fsbroken$ must be true, and therefore $A\aFS=B\aFS$ still holds.
\end{proof}

@@

We methodically investigate all possibilities using a computer 
program\footnote{The program is accessible on-line at \\
https://github.com/csirmaz/AlgebraicSyncPaper/blob/master/p2/prove.py.}
to determine
which pairs of commands cause errors all the time,
which can be simplified to one command or $\emptyseq$,
and which commute and can be reversed without any change in their overall effect.
Our basis for this investigation was the model of filesystems introduced in this paper.
These properties of command pairs are crucial as they determine
how a set of commands can be re-ordered to be applied to a filesystem
during synchronization, and what commands will never be compatible.
Below we list a number of statements derived using this method,
which are in fact lemmas that can be derived from our definitions of filesystems and of commands.

@@

\begin{mydef}[Assertion command]
A command is an assertion command if
it leaves every filesystem in the same state
where it is defined.
In other words, assertion commands have the same input and output type,
and there must also be only one possible value for their output.
Accordingly, $\cbba{n}$ and $\cdda{n}$ are the only two types of assertion commands.
\end{mydef}

@@

\begin{mylem}\label{incomparable_is_independent}
Two different commands that are not assertion commands are independent iff the nodes they change are incomparable:
\[ \cxynv\indep\czwmv \Longleftrightarrow n\unrel m \]
if $\cxynv\neq\czwmv$ and none of $\cxy$ and $\cxw$ is $\cbb$ or $\cdd$.
\end{mylem}
\begin{proof}
This is easy to show based on the output of the program that investigates the behavior of command pairs.
However, this proposition can also be derived from the \namecrefs{ax_separate_commute} already listed
in the following way.

We proceed in two steps. First, we note that
from \cref{ax_separate_commute,ax_separate_nobreaks} we know that
if $n\unrel m$, then $\cxynv$ and $\czwmv$ must necessarily be independent.

We prove the second step by contradiction.
We assume that both $\cxynv\indep\czwmv$ and $n\nunrel m$,
and show that it leads to contradiction.
As $n\nunrel m$
we know that $n=m$, $n\descendant m$ or $m\descendant n$.

If $n=m$, then from \cref{ax_same_breaks} we know $[y]=[z]$ and $[w]=[x]$
as otherwise the commands, in one order or the other, would break all filesystems
and they could not be independent.
The commands are not assertion commands, so from \cref{ax_same_singlec}
we know that $\cxynv\cc\czwnv\equiv\czwnv\cc\cxynv$ can only happen if $y=w$.
Therefore $[x]=[y]=[z]=[w]$ which with $y=w$ contradicts our assumption that the commands are different.

### 2016-10-15 p2/def-filesystem.tex
Starting with the definition of domains
Improved text

\begin{mydef}[Equivalence by type: $\typeeq$]
We write $v_1\typeeq v_2$ for $v_1,v_2\in\setv$ if $v_1$ and $v_2$ are both files,
are both directories, or are both $\empt$.
% TODO needed? v
In other words, if $\setf$ is the set of file values and $\setd$ is the set of directories, then
with $\{\empt\}$ they form a partition of $\setv$:
\[ \setv/{\typeeq} = \{\setd,\setf,\{\empt\}\}. \]
\end{mydef}

As usual, we write $[v]$ for the equivalence class of $v$, which is its type.
We also regard all file contents including all metadata to be included in the values in $\setf$.

The formal definition of the tree property is therefore
\begin{mydef}[Tree property]
\[ \forall n\in\setn: \FS(n) \neq \empt \Rightarrow \FS(\parent(n)) \in \setd \]
where $\parent(n)$ is defined.
\end{mydef}

@@

In this paper, $\FS$ and $\GS$ denote filesystems,
and $n$, $m$ and $o$ are nodes in $\setn$.
We will write $\valf$ for an arbitrary element in $\setf$.

>>>>

\begin{mydef}[Equivalence by type: $\typeeq$]
We write $v_1\typeeq v_2$ for $v_1,v_2\in\setv$ if $v_1$ and $v_2$ are both files,
are both directories, or are both $\empt$.
% In other words, if $\setf$ is the set of file values and $\setd$ is the set of directories, then
% with $\{\empt\}$ they form a partition of $\setv$:
% \[ \setv/{\typeeq} = \{\setd,\setf,\{\empt\}\}. \]
\end{mydef}

As usual, we write $[v]$ for the equivalence class of $v$, which represents its type.
We also consider metadata to be part of the values in $\setv$.

The formal definition of the tree property is therefore
\begin{mydef}[Tree property]
\[ \forall n\in\setn, \FS\in\setfs: \FS(n) \neq \empt \Longrightarrow \FS(\parent(n)) \in \setd \]
where $\parent(n)$ is defined, and where $\setd\subset\setv$ is the set of directories.
\end{mydef}

@@

In this paper, $\FS$ and $\GS$ denote filesystems,
and $n$, $m$ and $o$ are nodes in $\setn$.
We write $\valf$ for an arbitrary file value. % element in $\setf$.

### 2016-10-15 p2/rec-intro.tex
Improved text

\noindent
Next, we would now like to prove that $\reca$ and $\recb$ can be applied to the replicas,
that is, without loss of generality,
\[ \reca\aFS_B\neq\fsbroken. \]
So that we can formalize statements needed to prove this,
let us introduce two relations that describe under what conditions
sequences of commands work, that is, do not break a filesystem.

>>>>

\noindent
Next, we would now like to prove that $\reca$ and $\recb$ can be applied to the replicas,
that is, without loss of generality,
\[ \reca\aFS_B\neq\fsbroken. \]
So that we can formalize statements needed to prove this,
let us introduce relations that describe under what conditions
sequences of commands work, that is, do not break a filesystem.

### 2016-10-15 p2/works-inverse.tex
Improved text

\begin{mydef}[Sets of sequences]
As we will frequently refer to sets of fequences,
we will use calligraphic letters (e.g. $\seqset{A, B, C}$ and $\seqset{S}$)
to denote such sets for brevity.
We will use $A\cc\seqset{B}$ to denote the set of sequences $\{A\cc B|B\in\seqset{B}\}$,
and $\Dom{\sqs{A}}$ to denote $\bigcap_{A\in\sqs{A}} \Dom{A}$.
\end{mydef}

@@

As a special case, we write $\worksmnil{\sqs{A}}$ if we know that there is a filesystem
all sequences in $\sqs{A}$ work on,
and we apply these relations to sequences directly:
$\worksm{A}{B}$ with the meaning $\worksm{\{A\}}{\{B\}}$.
\end{mydef}

@@

\begin{myax}\label{combine_independent_commands}
The combination of independent commands works wherever the original commands work:
\[ \alpha\indep \beta \Rightarrow \worksm{\{\alpha, \beta\}}{\{\alpha\cc \beta\}}. \]
\end{myax}
\begin{proof}
We name this proposition a \namecref{combine_independent_commands} 
because to prove it, we must reach back to our filesystem model.
Let $\alpha=\cxynv$ and $\beta=\czwmv$.
%% TODO by contradiction
We proceed in an indirect way and
assume that there is a filesystem $\FS$ for which
$\cxynv\aFS\neq\fsbroken$ and $\czwmv\aFS\neq\fsbroken$, but
$\cxynv\cc \czwmv\aFS=\fsbroken$.
We know $\cxynv\aFS\neq\fsbroken$ so it must be applying 
$\czwmv$ that breaks it.
Applying a command can only result in a broken filesystem in three cases.
First, if the filesystem was already broken, which cannot be the case here.
Second, if the input type does not match the filesystem,
but we know $[\FS(m)]=[w]$ and so
$[(\cxynv\aFS)(m)]=[w]$ as based on \cref{incomparable_is_independent}, $n\neq m$.
Third, if the new filesystem violates the tree property.
This again cannot be the case because we also know that $n\unrel m$
and the tree property only depends on the types of the parent and children of $m$,
which therefore cannot be changed by $\cxynv$.
\end{proof}

@@

From \cref{incomparable_is_independent,ax_separate_commute} we know that
the commands in $S$ and $T$ pairwise commute, and so any sequence
that contains the commands from $S$ and $T$ and preserve their original partial order
is equivalent to $S\cc T$ on all filesystems.

@@

\begin{mylem}\label{worksinputmatch}
If $A$ and $B$ are minimal sequences, $\worksmnil{\{A,B\}}$,
and there are commands on node $n$ in both $A$ ($\cxynv\in A$) and $B$ ($\czwnv\in B$),
then the input types of these commands must match ($[x]=[z]$).
\end{mylem}
\begin{proof}
This result is similar to \cref{equiv_simple_same_commands} and
%% TODO by contradiction
is easily shown using an indirect proof: if $[x]\neq [z]$, then there is no filesystem that
either $\cxynv$ or $\czwnv$ would not break, 
and consequently $A$ and $B$ cannot work on the same filesystem.
\end{proof}

@@

%% TODO https://en.wikipedia.org/wiki/Equivalence_class

@@

As whether a sequence of commands breaks a filesystem does not depend on the output values
of the commands, only the input and output types, we know that
\begin{mycor}\label{repr_works_is_same}
\begin{gather*}
\works{\seqset{A}} \Leftrightarrow \works{\seqset{A}\T} \\
\worksc{\seqset{A}}{\seqset{B}} \Leftrightarrow \worksc{\seqset{A}\T}{\seqset{B}\T}.
\end{gather*}
\end{mycor}

@@

\begin{mylem}\label{repr_comm_inject}
Representative commands $\alpha\T$ are injections over representative filesystems ($\FS\T$)
except for the cases where they break a filesystem, that is,
\[ \forall\FS,\GS,\alpha: \FS\T\neq\GS\T \wedge \alpha\T(\FS\T)\neq\fsbroken \Longrightarrow \alpha\T(\FS\T) \neq \alpha\T(\GS\T). \]
\end{mylem}
\begin{proof}
This is trivial if $\GS\T=\fsbroken$, so we assume it is not so.
Let $\alpha\T = \cTxynv$.
If $\FS\T$ and $\GS\T$ have different values at any node apart from $n$, then this difference remains
even after applying $\cTxynv$ to $\FS\T$. The only remaining case therefore is where they have the same values
at all nodes except $n$.
As $\cTxynv(\FS\T)\neq\fsbroken$, we know $[\FS\T(n)]=[x]$.
We also know that $\GS\T(n)\neq\FS\T(n)$, and therefore, as there is a one-to-one correspondence between values and types,
$[\GS\T(n)]\neq[x]$.
This implies $\cTxynv(\GS\T)=\fsbroken$,
and again we have $\alpha\T(\FS\T) \neq \alpha\T(\GS\T)$.
\end{proof}

>>>>

\begin{mydef}[Sets of sequences]
As we will frequently refer to sets of fequences,
we will use calligraphic letters (e.g. $\sqs{A}$, $\sqs{B}$, $\sqs{C}$ and $\seqset{S}$)
to denote such sets for brevity.
We will use $A\cc\seqset{B}$ to denote the set of sequences $\{A\cc B|B\in\seqset{B}\}$,
and $\Dom{\sqs{A}}$ to denote $\bigcap_{A\in\sqs{A}} \Dom{A}$.
\end{mydef}

@@

As a special case, we write $\worksmnil{\sqs{A}}$ if we know that there is a filesystem
all sequences in $\sqs{A}$ work on.
We also apply these relations to sequences directly:
$\worksm{A}{B}$ with the meaning $\worksm{\{A\}}{\{B\}}$.
\end{mydef}

@@

\begin{myax}\label{combine_independent_commands}
The combination of independent commands works wherever the original commands work:
\[ \alpha\indep \beta \Rightarrow \worksm{\{\alpha, \beta\}}{\{\alpha\cc \beta\}}. \]
\end{myax}
\begin{proof}
We name this proposition a \namecref{combine_independent_commands} 
because to prove it, we must reach back to our filesystem model.
Let $\alpha=\cxynv$ and $\beta=\czwmv$.
We prove the proposition by contradiction and
assume that there is a filesystem $\FS$ for which
$\cxynv\aFS\neq\fsbroken$ and $\czwmv\aFS\neq\fsbroken$, but
$\cxynv\cc \czwmv\aFS=\fsbroken$.
We know $\cxynv\aFS\neq\fsbroken$ so it must be applying 
$\czwmv$ that breaks it.
Applying a command can only result in a broken filesystem in two cases.
First, if the input type does not match the filesystem,
but we know $[\FS(m)]=[w]$ and so
$[(\cxynv\aFS)(m)]=[w]$ as based on \cref{incomparable_is_independent}, $n\neq m$.
Second, if the new filesystem violates the tree property.
This again cannot be the case because we also know that $n\unrel m$
and the tree property only depends on the types of values at the parent and children of $m$,
which therefore cannot be changed by $\cxynv$.
\end{proof}

@@

From \cref{def_indep} we know that
the commands in $S$ and $T$ pairwise commute, and so any sequence
that contains the commands from $S$ and $T$ and preserve their original partial order
is equivalent to $S\cc T$ on all filesystems.

@@

\begin{mylem}\label{worksinputmatch}
If $A$ and $B$ are minimal sequences, $\worksmnil{\{A,B\}}$,
and there are commands on node $n$ in both $A$ ($\cxynv\in A$) and $B$ ($\czwnv\in B$),
then the input types of these commands must match ($[x]=[z]$).
\end{mylem}
\begin{proof}
This result is similar to \cref{equiv_simple_same_commands} and
is easily shown using a proof by contradiction: if $[x]\neq [z]$, then there is no filesystem that
either $\cxynv$ or $\czwnv$ would not break, 
and consequently $A$ and $B$ cannot work on the same filesystem.
\end{proof}

@@

@@

% TODO repeated
As whether a sequence of commands breaks a filesystem does not depend on the output values
of the commands, only the input and output types, we know that
\begin{mycor}\label{repr_works_is_same}
\begin{gather*}
\works{\seqset{A}} \Leftrightarrow \works{\seqset{A}\T} \\
\worksc{\seqset{A}}{\seqset{B}} \Leftrightarrow \worksc{\seqset{A}\T}{\seqset{B}\T}.
\end{gather*}
\end{mycor}

@@

% TODO partially bijective (?)
\begin{mylem}\label{repr_comm_inject}
Representative commands $\alpha\T$ are injections over representative filesystems ($\FS\T$)
except for the cases where they break a filesystem, that is,
\[ \forall\FS,\GS,\alpha: \FS\T\neq\GS\T \wedge \alpha\T(\FS\T)\neq\fsbroken \Longrightarrow \alpha\T(\FS\T) \neq \alpha\T(\GS\T). \]
\end{mylem}
\begin{proof}
This is trivial if $\GS\T=\fsbroken$, so we assume it is not so.
Let $\alpha\T = \cTxynv$.
If $\FS\T$ and $\GS\T$ have different values at any node apart from $n$, then this difference remains
even after applying $\cTxynv$ to $\FS\T$. The only remaining case therefore is where they have the same values
at all nodes except $n$.
As $\cTxynv(\FS\T)\neq\fsbroken$, we know $[\FS\T(n)]=[x]$.
We also know that $\GS\T(n)\neq\FS\T(n)$, and therefore, as there is a one-to-one correspondence between values and types,
$[\GS\T(n)]\neq[x]$.
This implies $\cTxynv(\GS\T)=\fsbroken$,
and again we have $\alpha\T(\FS\T) \neq \alpha\T(\GS\T)$.
\end{proof}
